Clustering, profiling and plots

### 2024-06-07 data/03-clustering-and-profiling.qmd
Small change (reproducibility) to the clustering-and-profiling.qmd

Select data for classification:
```{r}
df <- access_15_wide |>
  select(-c(cluster,
            starts_with("prop"), 
            starts_with("norm"))) |>
  rename(Dependent = total_Dependent_centric,
         Grocery = total_Grocery_centric,
         Errand = total_Errand_centric,
         Health = total_Health_centric) |>
  mutate(scluster = factor(scluster),
         div = round(div, digits = 3))

@@

Select data for classification:
```{r}
df <- access_15_wide |>
  transmute(scluster = factor(scluster),
            Income_bottom = `v_CA21_1103: In bottom half of the distribution`/(`v_CA21_1103: In bottom half of the distribution` + `v_CA21_1121: In top half of the distribution`),
            LICO = `v_CA21_1085: Prevalence of low income based on the Low-income cut-offs, after tax (LICO-AT) (%)`,
            Avg_numb_of_kids = `v_CA21_498: Average number of children in census families with children`,
            Not_in_labforce = `v_CA21_6504: Not in the labour force`/(`v_CA21_6498: Employed` + `v_CA21_6501: Unemployed` + `v_CA21_6504: Not in the labour force`),
            Income_gini = `v_CA21_1142: Gini index on adjusted household after-tax income`,
            Vis_minority = `v_CA21_4875: Total visible minority population`/ (`v_CA21_4914: Not a visible minority`+ `v_CA21_4875: Total visible minority population`),
            Age_0to14 =(`v_CA21_11: 0 to 14 years`) / (`v_CA21_11: 0 to 14 years` + `v_CA21_68: 15 to 64 years` + `v_CA21_251: 65 years and over`),
            One_parent_families = `v_CA21_507: Total one-parent families` / (`v_CA21_507: Total one-parent families` + `v_CA21_500: Total couple families`),
            walk_towork =  `v_CA21_7647: Walked` / (`v_CA21_7647: Walked` + `v_CA21_7635: Car, truck or van` + `v_CA21_7644: Public transit` + `v_CA21_7650: Bicycle` + `v_CA21_7653: Other method` ),
            Owner_hld_CHN = `v_CA21_4308: % in core housing need (57)`,
            Hld_ownership = `v_CA21_4306: % of owner households with a mortgage (58)`,
            Tenant_subsi_hld = `v_CA21_4314: % of tenant households in subsidized housing (61)`,
            Tenant_hld_CHH =`v_CA21_4316: % in core housing need (57)`,
            Population,
            Low_or_no_deg = (`v_CA21_5868: No certificate, diploma or degree` + `v_CA21_5871: High (secondary) school diploma or equivalency certificate`) / (`v_CA21_5868: No certificate, diploma or degree` + `v_CA21_5871: High (secondary) school diploma or equivalency certificate` + `v_CA21_5874: Postsecondary certificate, diploma or degree`)
            )

@@

#Train a decision tree using the data:
classdemo.som <- rpart(data = df,
                      formula = scluster ~ (Income_bottom + Age_0to14 + One_parent_families + Vis_minority + Not_in_labforce+ Income_gini+walk_towork+Owner_hld_CHN+Hld_ownership+Tenant_subsi_hld+Tenant_hld_CHH),
                      weights = Population
                      )
# # Visualize the decision tree:
# rpart.plot(classdemo.som, cex = 0.5, type = 0, varlen = 8)
```

>>>>

Select data for classification:
```{r}

df <- access_15_wide |>
  select(-c(cluster,
            starts_with("prop"), 
            starts_with("norm"))) |>
  rename(Dependent = total_Dependent_centric,
         Grocery = total_Grocery_centric,
         Errand = total_Errand_centric,
         Health = total_Health_centric) |>
  mutate(scluster = factor(scluster),
         div = round(div, digits = 3))

@@

Select data for classification:
```{r}

df <- access_15_wide |> left_join(access_15_aggregated, by = "from_id") |> 
  transmute(scluster = factor(scluster),
            Income_bottom = `v_CA21_1103: In bottom half of the distribution`/(`v_CA21_1103: In bottom half of the distribution` + `v_CA21_1121: In top half of the distribution`),
            LICO = `v_CA21_1085: Prevalence of low income based on the Low-income cut-offs, after tax (LICO-AT) (%)`,
            Avg_numb_of_kids = `v_CA21_498: Average number of children in census families with children`,
            Not_in_labforce = `v_CA21_6504: Not in the labour force`/(`v_CA21_6498: Employed` + `v_CA21_6501: Unemployed` + `v_CA21_6504: Not in the labour force`),
            Income_gini = `v_CA21_1142: Gini index on adjusted household after-tax income`,
            Vis_minority = `v_CA21_4875: Total visible minority population`/ (`v_CA21_4914: Not a visible minority`+ `v_CA21_4875: Total visible minority population`),
            Age_0to14 =(`v_CA21_11: 0 to 14 years`) / (`v_CA21_11: 0 to 14 years` + `v_CA21_68: 15 to 64 years` + `v_CA21_251: 65 years and over`),
            One_parent_families = `v_CA21_507: Total one-parent families` / (`v_CA21_507: Total one-parent families` + `v_CA21_500: Total couple families`),
            walk_towork =  `v_CA21_7647: Walked` / (`v_CA21_7647: Walked` + `v_CA21_7635: Car, truck or van` + `v_CA21_7644: Public transit` + `v_CA21_7650: Bicycle` + `v_CA21_7653: Other method` ),
            Owner_hld_CHN = `v_CA21_4308: % in core housing need (57)`,
            Hld_ownership = `v_CA21_4306: % of owner households with a mortgage (58)`,
            Tenant_subsi_hld = `v_CA21_4314: % of tenant households in subsidized housing (61)`,
            Tenant_hld_CHH =`v_CA21_4316: % in core housing need (57)`,
            Population,
            Low_or_no_deg = (`v_CA21_5868: No certificate, diploma or degree` + `v_CA21_5871: High (secondary) school diploma or equivalency certificate`) / (`v_CA21_5868: No certificate, diploma or degree` + `v_CA21_5871: High (secondary) school diploma or equivalency certificate` + `v_CA21_5874: Postsecondary certificate, diploma or degree`))

@@

#Train a decision tree using the data:
classdemo.som <- rpart(data = df,
                      formula = scluster ~ (Income_bottom + Age_0to14 + One_parent_families + Vis_minority + Not_in_labforce+ Income_gini+walk_towork+Owner_hld_CHN+Hld_ownership+Tenant_subsi_hld+Tenant_hld_CHH+LICO),
                      weights = Population
                      )
# # Visualize the decision tree:
# rpart.plot(classdemo.som, cex = 0.5, type = 0, varlen = 8)
```
